---
title: "Probabilistic multi-group comparison of alpha diversity"
author: "Rasmus Hindström"
date: last-modified
format:
    gfm:
        toc: true
---

# 0. Summary

This report demonstrates the use and interpretation of a probabilisitc alternative
to multi-group comparisons of alpha diversity. Group means are estimated using
a linear model fit with the `brms` package.


# 1. Data preparation

Preparing data from the `mia` package.

```{r}
#| label: load-libraries
#| code-summary: Load required libraries
#| echo: true
#| output: false
library(mia)
library(dplyr)
library(brms)
library(bayesplot)
library(dunn.test)
library(ggplot2)
library(patchwork)
library(rstatix)
library(microbenchmark)
```

```{r}
#| label: prepare-data
#| code-summary: prepare example data
data("peerj13075", package = "mia")
tse <- peerj13075
tse <- addAlpha(
    tse,
    assay.type = "counts",
    index = "shannon"
)
df <- as.data.frame(colData(tse))
```

# 2. Model fitting

The model is fit using the `brm` function from the `brms` package.
The response variable is the Shannon index, and the grouping variable is the 3 classes
of `age`; `Adult`, `Middle_age`, and `Elderly`.

The model is parametrized to model the group means directly, with no intercept.
Groups are allowed unequal variance, and means are modeled from a t distrubtion.
The shape parameter $\nu$ is estimated from data.

Model definition is as follows:

$$
y_{ik} \sim \text{t}(\nu, \mu_{ik}, \sigma_k)
$$

$$
\mu_{ik} = \beta_k
$$

$$
\sigma_k = \gamma_k
$$

*Default priors used by `brm()`*

$$
\nu \sim \gamma(2, 0.1)
$$
$$
\beta_0 \sim \text{t}(3, 1.3, 2.5)
$$
$$
\gamma_0 \sim \text{t}(3, 0, 2.5)
$$

$$
\beta_k, \gamma_k \sim \mathrm{Uniform}
$$


```{r}
#| label: fit-model
#| code-summary: Fit Bayesian model
#| warning: false
#| output: false 

# Model with no partial pooling 
fit <- brm(
    formula = bf(
        shannon ~ 0 + Age,
        sigma ~ 0 + Age
    ),
    data = df,
    family = student(),
    #control = list(adapt_delta = 0.9), # If divergence issues
    algorithm = "sampling",
    iter = 4000 
)
```

```{r}
#| label: fit-summary
#| echo: false
#| output: true
    
summary(fit)
```

Interpreting the coefficients and 95% CI we can already make the observation
that the groups appear to have rather different means.

Further plotting is required to make conclusions on other pair wise comparisons.

# 3. Posterior plotting

```{r}
#| label: plotting-post
#| code-summary: Posterior plotting
#| code-fold: true
#| fig-height: 6
#| fig-width: 12  
  
draws <- as_draws_df(fit)
population <- c(draws$b_AgeAdult, draws$b_AgeMiddle_age, draws$b_AgeElderly)
pop_mean <- mean(population)

plot_data <- data.frame(
    pop_mean = pop_mean,
    adult = draws$b_AgeAdult,
    adult_sd = draws$b_sigma_AgeAdult,
    elderly = draws$b_AgeElderly,
    elderly_sd = draws$b_sigma_AgeElderly,
    middle_age = draws$b_AgeMiddle_age,
    middle_age_sd = draws$b_sigma_AgeMiddle_age
)

p1 <- ggplot(data = plot_data) +
    geom_density(aes(x = adult), fill = "blue", alpha = 0.5, color = "blue") +
    geom_density(aes(x = middle_age), fill = "orange", alpha = 0.6, color = "orange") +
    geom_density(aes(x = elderly), fill = "purple", alpha = 0.7, color = "purple") +
    geom_vline(xintercept = plot_data$pop_mean, linetype = "dashed", color = "red", linewidth = 1) +
    labs(
        title = "Posterior Distributions of Group means",
        x = "Shannon index",
        y = "Density"
    ) +
    annotate(
        "text", x = 2, y = 2.5,
        label = "Blue: Adult\nOrange: Middle age\nPurple: Elderly"
    ) +
    theme_minimal()

p2 <- ggplot(data = plot_data) +
    geom_boxplot(aes(y = adult, x = 1), fill = "blue", alpha = 0.7, color = "black") +
    geom_boxplot(aes(y = middle_age, x = 2), fill = "orange", alpha = 0.7, color = "black") +
    geom_boxplot(aes(y = elderly, x = 3), fill = "purple", alpha = 0.7, color = "black") +
    geom_hline(yintercept = plot_data$pop_mean, linetype = "dashed", color = "red", linewidth = 1) +
    scale_x_continuous(
        breaks = c(1, 2, 3),
        labels = c("Adult", "Middle Age", "Elderly")
    ) +
    labs(
        title = "Boxplots of Group Means",
        x = "Group",
        y = "Shannon Index"
    ) +
    theme_minimal()

p1 + p2
```

From the plots we can infer groups are not similar. Particularly the 
Middle aged (Orange) group appears to have a lower Shannon index. The boxplots 
paint a clear picture of the higher overlap between the Adult and Edlerly group,
while the Middle aged group differs. In both plots the red dashed line indicates 
the total population posterior mean.

Using the posterior distributions, we can make statements about the differences
between the groups. In this context the probability of observing a lower Shannon index
is appropriate, akin to a classical p-value.

# 4. Quantify probabilities and effect sizes

```{r}
#| label: quantify-probabilities
#| code-fold: true
#| code-summary: Probabilities and Standardized Effect Size

calc_cohenD_post <- function(mu_1, mu_2, sd_1, sd_2) {
    # Calculates effect size from posterior draws
    diff <- mu_1 - mu_2
    pooled_sd <- sqrt((sd_1^2 + sd_2^2) / 2)
    d <- diff / pooled_sd
    mean_d <- mean(d)
    ci_d <- quantile(d, probs = c(0.05, 0.95))
    res <- list(
        "d" = mean_d,
        "ci" = ci_d
    )
    return(res)
}

calc_logfc_post <- function(mu_1, mu_2) {
    # Calculates log fold change from posterior draws
    logfc <- log2(mu_1 / mu_2)
    mean_logfc <- mean(logfc)
    ci_logfc <- quantile(logfc, probs = c(0.05, 0.95))
    res <- list(
        "logfc" = mean_logfc,
        "ci" = ci_logfc
    )
    return(res)
}

# Calculate effect sizes
logfc_adult_elderly <- calc_logfc_post(plot_data$adult, plot_data$elderly)
logfc_adult_middleage <- calc_logfc_post(plot_data$adult, plot_data$middle_age)
logfc_elderly_middleage <- calc_logfc_post(plot_data$elderly, plot_data$middle_age)

cohenD_adult_elderly <- calc_cohenD_post(
    plot_data$adult, 
    plot_data$elderly, 
    plot_data$adult_sd, 
    plot_data$elderly_sd
)
cohenD_adult_middleage <- calc_cohenD_post(
    plot_data$adult, 
    plot_data$middle_age, 
    plot_data$adult_sd, 
    plot_data$middle_age_sd
)
cohenD_elderly_middleage <- calc_cohenD_post(
    plot_data$elderly, 
    plot_data$middle_age, 
    plot_data$elderly_sd, 
    plot_data$middle_age_sd
)



probabilities <- data.frame(
    Comparison = c(
        "Adult vs Elderly",
        "Adult vs Middle age",
        "Elderly vs Middle age"
    ),
    Prob_lesser = c(
        prob_adult_elderly <- mean(plot_data$adult < plot_data$elderly),
        prob_adult_middleage <- mean(plot_data$adult < plot_data$middle_age),
        prob_elderly_middleage <- mean(plot_data$elderly < plot_data$middle_age)
    ),
    LogFC = c(
        logfc_adult_elderly$logfc,
        logfc_adult_middleage$logfc,
        logfc_elderly_middleage$logfc
    ),
    LogFC_ci_lower = c(
        logfc_adult_elderly$ci[1],
        logfc_adult_middleage$ci[1],
        logfc_elderly_middleage$ci[1]
    ),
    LogFC_ci_upper = c(
        logfc_adult_elderly$ci[2],
        logfc_adult_middleage$ci[2],
        logfc_elderly_middleage$ci[2]
    ),
    cohens_d = c(
        cohenD_adult_elderly$d,
        cohenD_adult_middleage$d,
        cohenD_elderly_middleage$d
    ),
    d_ci_lower = c(
        cohenD_adult_elderly$ci[1],
        cohenD_adult_middleage$ci[1],
        cohenD_elderly_middleage$ci[1]
    ),
    d_ci_upper = c(
        cohenD_adult_elderly$ci[2],
        cohenD_adult_middleage$ci[2],
        cohenD_elderly_middleage$ci[2]
    )
)

knitr::kable(probabilities, caption = "", format = "pipe")
```

Notice, that these are not classical p-values, but posterior probabilities.
The probabilities of observing a lower shannon index in the first group of each
comparison are reported. They have been calculated from the full posterior distribution.

Effect size's are reported as Log Fold Change (LogFC) and standardized effect size (Cohen's d) with 95% CI's.

Combining the probabilities and effect sizes, we can conclude that the
`Middle age` group has a lower Shannon index compared to the `Adult` group. 
With high probability (0.99) of observing a higher Shannon index in the `Adult` group
and a higher then 95% chance of observing an positive effect size.

The `Elderly` group has a similar pattern, with a high probabilty of observing a higher Shannon index than the `Middle age` group, but effect size 95% CI's overlap zero. This indicates that the difference is not as pronounced as with the `Adult` group. 

Comparing the `Adult` and `Elderly` groups, the probability of observing a higher Shannon index in the `Adult` group is greater then 50%, but the effect size is small, with a 95% CI that overlaps zero. This indicates that the difference is not likely to be meaningful.

# 5. Classical approachs to multi-group testing

## 5.1. ANOVA

ANOVA would be the closest classical alternative. Assumptions in ANOVA are
normality and equal variance. 

```{r}
#| label: anova
res_anova <- aov(shannon ~ Age, data = df)

summary(res_anova)

```

Our result inidcates that Age is a significant predictor on explaining differences
in the Shannon index. After observing a significant result, it is typical to further
inspect the data with a pairwise t-test. Another option is
Tukey's Honest Significant Difference (HSD). Both of which require adjusting p-values
due to being post-hoc tests.

```{r}
#| label: pairwise-t-test

# Pairwise t-test
pairwise.t.test(df$shannon, df$Age, p.adjust = "fdr")

# HSD
TukeyHSD(aov(shannon ~ Age, data = df))
``` 

Both post-hoc tests point to the significant difference between the groups
`Adult` and `Middle age`. In the pairwise t-test the difference between groups
`Middle age` and `Elderly` approaches the significance boundry (p < 0.05).

## 5.2. Kruskal-Wallis

Another option to ANOVA is the Kruskal-Wallis test. Kruskal-Wallis relaxes the
assumption of normality. However, It also needs to be paired with 
a post-hoc test to infer paired differences after global significance has been tested.
Kruskal-Wallis is typically paired with Dunn's post-hoc test.

```{r}
#| label: kruskal-test
kruskal.test(shannon ~ Age, df)
```

We get a p-value < 0.05, so there are 'significant' differences within the groups.

```{r}
#| label: DUNN's test
dunn.test(df$shannon, df$Age, method = "bh", kw = FALSE)
```

Significant p-values, after adjustment, are reported for the comparison between
groups `Adult` and `Middle age`.

# 6. Benchmarking speed

Using the `microbenchmark` package we can compare the raw speed of the methods.

```{r}
#| label: table-comparison
#| code-fold: true
#| code-summary: Comparison of run times

.bayes_est <- function(df) {
    fit <- brm(
    formula = bf(
        shannon ~ 0 + Age,
        sigma ~ 0 + Age
    ),
    data = df,
    family = student(),
    #control = list(adapt_delta = 0.9), # Can help with divergence issues
    algorithm = "sampling",
    iter = 4000
    ) 
    return(fit)
}

.anova <- function(df) {
    fit <- aov(shannon ~ Age, data = df)
    res_posthoc <- TukeyHSD(fit)
    return(
        list(
            fit = fit,
            post_hoc = res_posthoc
        )
    )
}

.kruskal_test <- function(df) {
    fit <- kruskal.test(shannon ~ Age, df)
    res_posthoc <- dunn.test(df$shannon, df$Age, method = "bh", kw = FALSE)
    return(
        list(
            fit = fit,
            post_hoc = res_posthoc
        )
    )
}

mb_res <- microbenchmark(
    .bayes_est(df),
    .anova(df),
    .kruskal_test(df),
    times = 1L, # adjust for more repetitions
    unit = "seconds"
)

mb_res
``` 

It is clear that the classical methods are orders of magnitude faster then the 
bayesian model. This highlights one of the main limitations of bayesian or probabilistic
methods, the need for more computational resources to run the sampler. In addition
the `brms` package generates and compiles the model into stan code. Using the `rstanarm`
package might offer speed benefits, due to it relying on precompiled stan models. 
However, this sacrifces the flexibility and user-friendly interface of `brms`. 

# 7. Conclusions

Although the methods are in aggreement that the significant outlier is
the `Middle age` group. The classical methods are only able to provide a binary (yes/no)
inference, in contrast to the much richer and intuitive information provided by the probabilistic
method. 

The main drawback of the probabilistic approach is the computational cost.
Fitting a Bayesian model, with brms, requires compiling the model code and running
the MCMC sampler. However this margin may shrink when datasets grow larger, as 
probabilistic model fitting scales well larger data.
