---
title: "Probabilistic workflow for univariate group comparison"
author: "Rasmus Hindström"
date: last-modified
format:
    gfm:
        toc: true
---

# 0. Introduction 

Having established that probabilisic methods can be used as alternatives to 
classical methods in univariate group comparison, we now prepare a recommended
workflow to follow. 

# 1. Install and load packages

```{r}
#| label: package-installation
#| eval: false
#| echo: true
#| output: false

packages <- c(
    "bayesplot",
    "brms",
    "dplyr",
    "ggplot2",
)

to_install <- packages[!packages %in% installed.packages()[, "Package"]]
if (length(to_install) > 0) {
  install.packages(to_install)
}

bioc_pkg <- "mia"

if (!requireNamespace(bioc_pkg, quietly = TRUE)) {
  BiocManager::install(bioc_pkg)
}
```

```{r}
#| label: load-libraries
#| echo: true
#| output: false

library(bayesplot)
library(brms)
library(dplyr)
library(ggplot2)
library(mia)
```

# 2. Prepare data

We work under the assumption that data is in a TSE -object. For more information on 
microbiome data and use of TSE -objects refer to
(Orchestrating Microbiome Analysis)[https://microbiome.github.io/OMA/docs/devel/index.html]

For demonstration purposes we show the workflow with an example dataset contained
in the `mia` -package.

```{r}
#| label: data-wrangling

data("peerj13075", package = "mia")  
tse <- peerj13075 

# You may need to agglomerate your data refer to OMA
#tse <- agglomerateByRank(tse, rank = "genus")

tse <- addAlpha(
    tse,
    assay.type = "counts",
    index = "shannon"
)

# Extracting the data into a dataframe for easier modeling
df <- as.data.frame(colData(tse))

str(df)
```

# 3. Fitting a probabilistic model

## 3.1. No pooling model

Without partial pooling we still get regularization from priors. The model is 
parametrized to estimate the group means directly for straight forward comparison.
This type of parametrization is also slightly more efficient to sample.

```{r}
#| label: no-pool-fit
#| output: false 

fit1 <- brm(
    formula = bf(
        shannon ~ 0 + Age,
        sigma ~ 0 + Age
    ),
    data = df,
    family = student(),
)
```

```{r}
#| label: no-pool-summary
#| code-fold: true

summary(fit1)  
```

## 3.2. Partial pooling model

Fitting a hierachical model with partial pooling, provides shrinkage to correct
for multiple testing more strongly. This type of model is harder to sample, and
requires some tuning to get the sampler to behave well. 

```{r}
#| label: pool-fit
#| output: false 

fit2 <- brm(
    formula = bf(
        shannon ~ 0 + (1 | Age),
        sigma ~ 0 + (1 | Age)
    ),
    data = df,
    family = student(),
)

fit2 <- update(
    fit2,
    control = list(adapt_delta = 0.9), # Tighter acceptance
    iter = 8000 # Longer chains
)
```

```{r}
#| label: partial-pooling-summary
#| code-fold: true 
 
summary(fit2)

# Random effects
ranef(fit2)
```

Notice that compiling the model takes some time, but refitting the model with 
`update()` is considerably faster. Switching to variational inference can bring
even larger speed benefits over MCMC, but you sacrifice accuracy of the posterior.

## 3.3. Model comparison

We can quickly compare model performance by aproximating out-of-sample preformance,
using PSIS-LOO. This is the equivalent to LOO-CV in the classical paradigm. 

```{r}
#| label: model-comparison

loo(fit1, fit2) 
```

Model performance is equivalent. `se_diff` values are larger then the `elpd_diff`, there is no indication of either model being better then the other on out-of-sample prediction. 

I would argue that for our purpose, the no-pooling model is approriate when
there are so few groups. The regularisation from the priors is enough, but when 
there are more groups compared the partial pooling model would be favored.

# 4. Plotting and inference

We will proceed with the no-pooling model. Extracting the posterior draws,
allows easy plotting to visualize differences, compute comparisons of interest,
and make probabilistic statements or inference.

## 4.1. Posterior plotting

```{r}
#| label: plotting
#| warning: false 
#| code-fold: true
#| code-summary: Plot code 
#| fig-height: 6
#| fig-width: 12
 
# Get the draws
draws <- as_draws_df(fit1)
draws <- draws %>%
    select(starts_with(paste0("b_", "Age")))
pop_mean <- mean(unlist(draws))

# Density plot
p1 <- ggplot(data = draws) +
    geom_density(aes(x = b_AgeAdult), fill = "blue", alpha = 0.5, color = "blue") +
    geom_density(aes(x = b_AgeMiddle_age), fill = "orange", alpha = 0.6, color = "orange") +
    geom_density(aes(x = b_AgeElderly), fill = "purple", alpha = 0.7, color = "purple") +
    geom_vline(xintercept = pop_mean, linetype = "dashed", color = "red", linewidth = 1) +
    labs(
        title = "Posterior Distributions of Group means",
        x = "Shannon index",
        y = "Density"
    ) +
    annotate(
        "text", x = 2, y = 2.5,
        label = "Blue: Adult\nOrange: Middle age\nPurple: Elderly"
    ) +
    theme_minimal()

# Box plot
p2 <- ggplot(data = draws) +
    geom_boxplot(aes(y = b_AgeAdult, x = 1), fill = "blue", alpha = 0.7, color = "black") +
    geom_boxplot(aes(y = b_AgeMiddle_age, x = 2), fill = "orange", alpha = 0.7, color = "black") +
    geom_boxplot(aes(y = b_AgeElderly, x = 3), fill = "purple", alpha = 0.7, color = "black") +
    geom_hline(yintercept = pop_mean, linetype = "dashed", color = "red", linewidth = 1) +
    scale_x_continuous(
        breaks = c(1, 2, 3),
        labels = c("Adult", "Middle Age", "Elderly")
    ) +
    labs(
        title = "Boxplots of Group Means",
        x = "Group",
        y = "Shannon Index"
    ) +
    theme_minimal()

p1 + p2
```

## 4.2. Probabilistic inference

```{r}
#| label: table

# Get pairwise comparisons
comparisons <- combn(names(draws), 2, simplify = FALSE)

summary_tbl <- lapply(comparisons, function(cols) {
  g1 <- cols[1]
  g2 <- cols[2]
  
  prob <- mean(draws[[g1]] > draws[[g2]])
  log2fc_vals <- log2(draws[[g1]] / draws[[g2]])
  
  tibble(
    comparison = paste(g1, "vs", g2),
    prob_g1_gt_g2 = prob,
    log2fc_mean = mean(log2fc_vals),
    log2fc_lower = quantile(log2fc_vals, 0.025),
    log2fc_upper = quantile(log2fc_vals, 0.975)
  )
}) %>% bind_rows

knitr::kable(summary_tbl, type = "pipe")
```

