---
title: "Validation of probabilistic group comparison"
author: "Rasmus Hindström"
date: last-modified
format:
    gfm:
        toc: true
---

# False positives

Taking a dataset, and randomly assigning groups should not produce meaningful
differences. Testing methods on this random assignment dataset should
not produce any significant results. If some are detected, that is a false
positive. 

```{r}
#| label: setup
#| output: false

library(mia)
library(bayestestR)
library(brms)
library(dplyr)
library(ggplot2)
library(ggsci)
library(tidybayes)  

options(mc.cores = 4)
```

```{r}
#| label: data-prep
#| code-fold: true
#| code-summary: Data preparation

data(Tito2024QMP, package = "mia")
tse <- Tito2024QMP

tse <- addAlpha(
    tse,
    assay.type = "counts",
    index = "shannon"
)

df <- as_tibble(colData(tse))

# Assign random group 
groups <- c("A", "B", "C", "D")
df <- df %>%
    mutate(group = sample(groups, n(), replace = TRUE))

# Check means and group size
df %>%
    group_by(group) %>%
    summarise(
        mean_shanonn = mean(shannon),
        N = n()
        )
```

## Fitting models

As a baseline lets first do ANOVA, we should not get a significant result.

```{r}
res_aov <- aov(shannon ~ group, data = df)
summary(res_aov)
```

Indeed, no global significance. Against theory, we still do a post-hoc test.

```{r}
TukeyHSD(res_aov)
```

Even when trying our best to get a FP we dont. 

Lets try fitting a probabilistic model with `brms`.

```{r}
# Worst case, no pooling
formula = bf(
    shannon ~ 0 + group,
    sigma ~ 0 + group
)

mod <- brm(
    formula = formula,
    data = df,
    family = gaussian(),
    iter = 2e3
)

# Posterior Predictive
pp_check(mod, ndraws = 50)
```

The no partial pooling model, only has regularization from priors. 
No shrinkage from a hierachical structure. 

```{r}
raw_means <- df %>%
    group_by(group) %>%
    summarise(mean = mean(shannon))
pop_mean <- df %>%
    summarise(mean = mean(shannon))

```
